A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-04 latex/main.tex
WIP for discussion

% \section*{DISCUSSION}

% NOTES: paragraph structure. Lead with the biological insights first 
% to prevent the ``this is a methods paper'' criticism

% \begin{itemize}
% \item We present a landmark ARG, which has a comprehensive representation
% of evolutionary history.
% \item This is a conservative lower-bound on the occurance of recombination.
% Very high validation rate post-QC.
% Substantially lower than previous estimate. 
% The first order term of this 
% is the choice of $p$ parameter. With p=4 we show that the ARG contains
% more of the previously identified recombinants.
% \item Truth data for recombination is difficult to find. The closest we have is the 
% pango X lineages. No methods agrees with all of the designations being 
% true recombinants. Our analysis suggests that there is significant over-naming
% with the same underlying recombination event getting associated with 
% multiple X lienages at least twice. Also many lineages are not suggested 
% as being recombinant. This resolves long-standing complexity in 
% the Omicron evolution and simplifies our understanding.
% \item Based on this cleaned up and conservative estimate of recombination
% prevalance. Supports the idea that  recombination frequency in SARS-CoV-2 
% is driven jointly by viral
% load in the population and lineage diversity, implying predictable
% recombination ``hot periods'' during pandemics
% \item Also found evidence that breakpoints are enriched near spike, 
% supporting findings of earlier studies.
% Suggests functional constraints or template-switching
% biases that may shape SARS-CoV-2 evolution.
% Spike gene junctions are recurrent recombination zones, likely under both
% mechanistic and selective pressures.
% \item This ARG is base on a new method, sc2ts. Major methodological 
% breakthrough in terms of handling recombination. We validate the method
% by using results from the literature. We don't use simulations because 
% there's no real point in this case. We're just focusing on SARS-CoV-2.
% We could create some simulations and generate some metrics about how
% well sc2ts picks up simulated recombination events on clean data, 
% but what would that tell us ultimately? (Similarly for tree reconstruction.)
% Certainly standardised simulations would help to refine methods, but
% we have moved the bar forwards massively here and made a great starting
% point for subsequent method development.
% \item There is great potential for using this ARG in future studies,
% and using sc2ts as a baseline to judge future methods.
% \item The tskit and VCF Zarr toolchains enable this research.
% \item We have shown that recombination can be handled, provided
% feasible methods to infer it, and shown that a preexisting tool
% chain can provide quick insignts. The future is bright.

% \end{itemize}

\section*{DISCUSSION}
[NOTE: this is two different WIP versions of the Discussion. 
The ChatGPT based stuff was only ever intended as a starting point,
and not to be taken literally.]

% P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% this has more.
% SHZ: This makes it sound like we are just making analyses
% (which people can do already using existing methods and tools)
% easier to do
% rather than making a leap forward from anything else out there.
We present the first pandemic-scale ancestral recombination graph (ARG) for
SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
ARG represents a step change in how the evolutionary history of this virus can
be represented and analysed. Like the UShER phylogeny, it provides a
comprehensive and continuously updatable record of global genomic diversity,
but it extends this paradigm by embedding recombination directly into the
genealogy. The ARG unifies point mutations, deletions, and recombination events
within a single, compact data structure, enabling analyses that
were previously fragmented across multiple methods and data models. By encoding
reticulate ancestry explicitly rather than assuming a single tree, the ARG
provides a complete and internally consistent representation of SARS-CoV-2
evolution. In this way, the
ARG offers the same accessibility and scalability that made the UShER tree
indispensable during the pandemic, while expanding its scope to encompass the
full spectrum of evolutionary processes shaping SARS-CoV-2.

% P2. This is a conservative lower-bound on recombination. 
% Discuss other estimates.
% 
% SHZ: Why should we care about getting this number/frequency right?
% Sure, popgen folks love estimating this, but to other people,
% existing estimates already tell us that
% detectable recombination in SARS-CoV-2 is (qualitatively) rare,
% even without the advances of sc2ts.
% Also, we are talking about the frequency of detectable recombination
% in the last sentence, right?
% The true frequencey of recombination is unknowable from this type of data.
% So, the best we are ever going to get is a better lower bound.
The recombination frequency estimated from the \texttt{sc2ts} ARG is
substantially lower than that implied by the combination of UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
reported 589 unique recombination events with 43,104 descendant
samples---corresponding to roughly 2.7\% of all genomes being of recombinant
origin.  Using the same time interval, our inference at \(k=4\) identifies only
14 high-confidence recombination events with 172 descendants among 501,235
genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
explained by the stricter parsimony penalty adopted here: increasing the
penalty from \(p=3\) (as used in the RIPPLES analysis) 
to \(k=4\) greatly reduces false
positives while retaining nearly all verified recombinant lineages, 
including those missed by RIPPLES at equivalent thresholds.  
We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
of recombination as
tree rearrangements proceed, thereby requiring the more lenient 
threshold in order to find known recombinats.
These comparisons suggest that our estimate represents a conservative 
lower bound. Further work will be required to determine
the true frequency of recombination in SARS-CoV-2 evolution.

% P3. Clarifies relationships between pango X lineages, simplifying
% the Omicron sublineage.
% 
% SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
% Sure, phylogeneticists would get a kick out of it,
% but I don't think the majority of Cell readership would.
% I think what is important is knowing
% how many emerging recombinants carrying new combinations of mutations of concern
% we are actually dealing with.
% Clarifying the relationships between Pango Xs helps with this task.

The Pango X lineage designations represent the community’s best attempt to
catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
manual, volunteer-driven process that synthesizes multiple lines of evidence
and expert judgment rather than a single automated criterion.  As a result, the
system provides the closest available approximation to “truth data,” but
inevitably contains inconsistencies and over-splitting.  Within the
\texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
directly to recombination nodes (Type I events).  Examination of their
genealogical relationships shows that many of these 44 lineages share common
ancestral recombination events rather than arising independently.  In
particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
and XAP—share a second common ancestor that we refer to as “Xx.”  These
clusters account for roughly half of all X designations in the dataset,
indicating that the community process has at times produced multiple labels for
the descendants of a single recombinant lineage.  Conversely, several [TODO 
be more concrete]
designated X lineages show little or no evidence of recombination in the ARG,
suggesting that some may reflect convergent mutation or limited sampling rather
than hybrid origin.  Together, these results provide a data-driven synthesis in
which the 44 currently recognized recombinant lineages collapse into a smaller
number of well-supported ancestral events, simplifying the complex picture of
Omicron-era evolution and bringing the community nomenclature into closer
alignment with the underlying genealogy.

% P4. Results on global case numbers and frequency of recomb.
% SHZ: Should we mention how the reliability of global case number data
% and variation in sampling effort over time and across countries
% might affect our interpretation of these findings?
[FIXME: This is direct output from ChatGPT.  Review!]
The dated ARG allows us to place recombination within its epidemiological
context, revealing clear temporal structure in when recombinant lineages arise.
The frequency of inferred recombination events follows global case numbers
through the pandemic, with peaks during the Delta and Omicron waves when both
infection prevalence and genetic diversity were high.  Quantitatively, the best
single predictor of weekly recombinant counts is the product of global cases
and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
relationship is consistent with the mechanistic expectation that recombination
requires both frequent co-infections (driven by high case numbers) and
genetically distinct co-infecting viruses (driven by lineage diversity).  In
other words, recombination becomes most detectable when epidemiological and
evolutionary opportunity coincide.  These results imply that the emergence of
novel recombinants is not uniform through time but concentrated into
predictable ``hot periods'' during large, antigenically diverse outbreaks, when
the ecological conditions for viral recombination are most favourable.

% P5. Confirm breakpoints around Spike. 
% SHZ: Maybe we can say it's a known (instead of robust)
% pattern of detectable recombination in CoVs in general?
% There are some citations for template switching hotspots in SC2.
% Also, it's not only about selection favoring new combinations of mutations in Spike.
% It's about combinations of mutations in, say, ORF1ab, getting selected against
% (or just low mutation rates).
% Maybe mention epistasis?
[FIXME: This is directish output from ChatGPT. Review!]
The genomic distribution of breakpoints in the ARG is highly uneven, with a
pronounced enrichment near the $5'$ boundary of the Spike gene and across the
S1–S2 junction.  This pattern closely mirrors that observed in the
UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
analyses \citep{Jackson2021} [more cites], but here it is recovered
independently over a much longer sampling period and under a more conservative
inference regime.  Under our more conservative estimate of the overall
prevalence of recombination, we find that this enrichment persists despite a substantial
reduction in the total number of inferred recombination events, indicating that
it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
by-product of detection sensitivity.  Across all high-confidence events, the
median breakpoint interval length is $\sim 2.2$kb and declines with parental
divergence, consistent with expectations from sequence-based localization
accuracy.  These results show that even with a conservative lower bound on
recombination frequency, the concentration of breakpoints around Spike remains
a robust signature of coronavirus genome biology—likely reflecting a
combination of mechanistic biases in template switching and selective
constraints on Spike function.

% P6. Sc2ts method is a major advance. We validate with real data
% rather than simulations. Forward looking statement about simulations.
% 
% SHZ: Could at least some of this go into the Introduction?
In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
it is now possible to reconstruct recombination-aware genealogies at the scale
of entire pandemics.  Although recombination is recognised as a major
evolutionary force in many pathogens—including RNA viruses such as influenza
and HIV, and numerous bacterial species—genealogical analyses have typically
relied on tree-based approximations that ignore or collapse recombination.
This simplification has limited the accuracy of downstream evolutionary and
epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
overcomes this barrier by directly inferring an ancestral recombination graph
from millions of viral genomes, producing the first coherent reticulate
genealogy for any pathogen dataset of comparable size.  The ability to
integrate recombination into genome-wide genealogies at this scale represents a
major conceptual step forward: it transforms recombination from a confounding
complication into an analyzable component of pathogen evolution.

% P7. The software ecosystem enables diverse besplike analysis.

[TODO: This is close, needs revision]
The practical value of the ARG is amplified by its integration into the broader
software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
open-source libraries provide highly efficient storage and analysis of tree
sequences and variant data, allowing millions of genomes to be explored
interactively within familiar environments such as Python and Jupyter
notebooks.  Analyses that previously required bespoke phylogenetic
pipelines—such as extracting local genealogies, summarising ancestral
relationships, mapping mutations to branches, or tracking recombinant ancestry
through time—can now be performed with concise, reproducible code on standard
hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
ARG integrates seamlessly with widely used tools in the PyData ecosystem,
enabling users to combine genealogical information with epidemiological,
geographic, or phenotypic metadata in the same workflow.  This interoperability
transforms what can be done with large viral genome collections, lowering the
barrier to sophisticated evolutionary analysis and making recombination-aware
genealogy inference accessible to a wide community of researchers.

% P8. Outlook

[TODO a bit muddled, refine]
Together, these results demonstrate that recombination in SARS-CoV-2 can be
systematically incorporated into genealogical inference at pandemic scale. The
\texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
both biologically realistic and computationally tractable, showing that
recombination can be modelled rather than ignored. Embedded within an
interoperable software ecosystem, it provides a comprehensive record of
SARS-CoV-2 ancestry and a practical framework for real-time,
recombination-aware genomic surveillance—turning large sequence archives into
dynamic evolutionary models that can be queried with concise, reproducible code. 
More broadly, this work marks a transition from
tree-based pandemic phylogenetics to recombination-aware genealogy inference.
While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
point for other recombining pathogens: the core ideas are portable,
even though effective deployment will require pathogen-specific adjustments. 
We anticipate that this resource will clarify the recent history of
SARS-CoV-2 and catalyse methodological refinement and targeted biological
studies of recombination across systems where such extensions prove feasible.

[NOTE: This is Shing's WIP alternative]

%%% SHZ: WIP alternative.
We present the first pandemic-scale
ancestral recombination graph (ARG) for SARS-CoV-2,
reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
We demonstrate how this ARG, for the first time,
enables systematic identification and characterization of
recombination events in fine-grained
detail--in terms of their parents, breakpoints, and timing of
emergence--providing richer and more nuanced
genealogical, genomic, and epidemiological context
than existing phylogenetic approaches.
Together, sc2ts and the ARG represent a new paradigm in
how we study the evolution of pathogens in the face of ongoing sampling.

Accounting for recombination is a key challenge in pathogen genomics, and
recombination has been largely ignored in analysis of viral genomes.
Existing phylogenetic approaches for pathogens detect recombination
\textit{post hoc} (e.g., UShER+RIPPLES),
treat it as a nuance (e.g., XX), or
discard its signals from downstream analyses (e.g., XX).
Sc2ts addresses this challenge
by using the ARG as the underlying data model,
explicitly representing recombination events as individual nodes and
relating them to specific parents in the ARG.
The ARG provides a coherent framework to determine and represent
the relationships between recombinants and their specific parents
(which are not possible using a tree-based data model, e.g., the MAT format of UShER).
The ARG reveals and clarifies the relationships between
individual samples of the same recombinant origin,
avoiding issues such as over-naming recombinants
(e.g., the samples assigned to
five differently labeled recombinant Pango lineages
descending from the “Xx” recombination event in the ARG).
Our results show how an ARG can considerably simplify
efforts to identify and monitor emerging recombinants,
facilitating communication of genomic surveillance results.

% SHZ: Very WIP
% Discuss the new preprint from Turakhia.
The timing of recombination events is of key interest in
pathogen genomic epidemiology.
because this information may yield insights into
how the emergence of a new recombinant might be
linked with changes in public health policies.

The overall frequency of detectable recombination events
estimated from the sc2ts ARG is substantially lower than
that from UShER+RIPPLES,
largely reflecting differences in model formulation and parameter choice.
When applied \textit{post hoc} to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
Using the same time interval, sc2ts \(k=4\) identifies
14 high-confidence recombination events with
172 descendant samples among 501,235 samples \(0.034\%\),
a 40- to 80-fold reduction.
Much of this difference is explained by
the stricter parsimony penalty used here:
increasing the penalty from p=3 (as used in the RIPPLES analysis) to
k=4 greatly reduces false positives
while retaining nearly all verified recombinant lineages,
including those missed by RIPPLES at equivalent thresholds.
These indirect comparisons suggest that
our estimate represents a conservative lower bound.
In future studies, dissecting the methodological differences
between sc2ts and UShER+RIPPLES may
inspire new ideas to develop or improve ARG building methods
for SARS-CoV-2 (and potentially other pathogens).

The ARG itself is a powerful resource
to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies
important features of SARS-CoV-2 evolution,
which are shown in different studies
using various methods, tools, and approaches.
The ARG also compactly encodes genetic variation in
millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of
this valuable dataset for a range of
genetic and evolutionary analyses beyond those presented here,
for instance, to understand
how the variability in sequencing efforts
across time and location affects our ability
to reconstruct the origin of important recombinant lineages and
to estimate when they emerge.
Because building the ARG is computationally expensive, and
checking and improving various aspects of its quality is laborious,
we hope that making this resource openly available will
save significant time and effort for other researchers,
enabling immediate focus on analyzing the information
contained within the COVID-19 dataset.

% TODO: Briefly discussion software tools here or in a subsequent paragraph.

Recent developments in ARG methods have focused almost exclusively on
applications to human sequencing data.
The availability of human ARGs reconstructed from large datasets
has already had a tremendous impact on improving
our understanding of human genetic ancestry and past demography.
We anticipate similar advances to be made
by adapting existing inference methods for human ARGs
to use the sc2ts ARG for the genealogy-based study of
natural selection or the geographic spread of SARS-CoV-2 through time.

The unprecedented throughput and pace of genome sequencing are
creating ample opportunity to understand the impacts of recombination
on pathogen evolution and epidemiology.
We hope that this study will spur
development of ARG methods for pathogens,
ultimately making recombination-aware genome analysis
routine for all sorts of pathogens beyond SARS-CoV-2.

@@

% SHZ: Should we do point form?
We close by emphasizing several current limitations
as well as potential directions to improve sc2ts,
which we hope will also inspire future research
to develop ARG methods for SARS-CoV-2 and other pathogens.

Reversion of deletions can indicate recombination,
because bases that are lost are unlikely to be regained,
except via recombination with a lineage that has those bases.
We presently treat multi-base deletions as missing data.
Modeling deletions (and more generally insertions) in the HMM
could help recover additional signals of recombination
and reduce false positives.

>>>>

\section*{DISCUSSION}
There are two primary contributions of this work. The first is a
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analysis widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
if we like this].
We hope that the ARG provided here will similarly
enable a range recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements fully integrating
recombination into pathogen genomics.

The second primary contribution of this work is the method used to 
infer the ARG, sc2ts. [WIP]

% Old discussions

% % P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% % this has more.
% % SHZ: This makes it sound like we are just making analyses
% % (which people can do already using existing methods and tools)
% % easier to do
% % rather than making a leap forward from anything else out there.
% We present the first pandemic-scale ancestral recombination graph (ARG) for
% SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
% ARG represents a step change in how the evolutionary history of this virus can
% be represented and analysed. Like the UShER phylogeny, it provides a
% comprehensive and continuously updatable record of global genomic diversity,
% but it extends this paradigm by embedding recombination directly into the
% genealogy. The ARG unifies point mutations, deletions, and recombination events
% within a single, compact data structure, enabling analyses that
% were previously fragmented across multiple methods and data models. By encoding
% reticulate ancestry explicitly rather than assuming a single tree, the ARG
% provides a complete and internally consistent representation of SARS-CoV-2
% evolution. In this way, the
% ARG offers the same accessibility and scalability that made the UShER tree
% indispensable during the pandemic, while expanding its scope to encompass the
% full spectrum of evolutionary processes shaping SARS-CoV-2.

% % P2. This is a conservative lower-bound on recombination. 
% % Discuss other estimates.
% % 
% % SHZ: Why should we care about getting this number/frequency right?
% % Sure, popgen folks love estimating this, but to other people,
% % existing estimates already tell us that
% % detectable recombination in SARS-CoV-2 is (qualitatively) rare,
% % even without the advances of sc2ts.
% % Also, we are talking about the frequency of detectable recombination
% % in the last sentence, right?
% % The true frequencey of recombination is unknowable from this type of data.
% % So, the best we are ever going to get is a better lower bound.
% The recombination frequency estimated from the \texttt{sc2ts} ARG is
% substantially lower than that implied by the combination of UShER+RIPPLES,
% largely reflecting differences in model formulation and parameter choice.
% RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
% reported 589 unique recombination events with 43,104 descendant
% samples---corresponding to roughly 2.7\% of all genomes being of recombinant
% origin.  Using the same time interval, our inference at \(k=4\) identifies only
% 14 high-confidence recombination events with 172 descendants among 501,235
% genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
% explained by the stricter parsimony penalty adopted here: increasing the
% penalty from \(p=3\) (as used in the RIPPLES analysis) 
% to \(k=4\) greatly reduces false
% positives while retaining nearly all verified recombinant lineages, 
% including those missed by RIPPLES at equivalent thresholds.  
% We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
% of recombination as
% tree rearrangements proceed, thereby requiring the more lenient 
% threshold in order to find known recombinats.
% These comparisons suggest that our estimate represents a conservative 
% lower bound. Further work will be required to determine
% the true frequency of recombination in SARS-CoV-2 evolution.

% % P3. Clarifies relationships between pango X lineages, simplifying
% % the Omicron sublineage.
% % 
% % SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
% % Sure, phylogeneticists would get a kick out of it,
% % but I don't think the majority of Cell readership would.
% % I think what is important is knowing
% % how many emerging recombinants carrying new combinations of mutations of concern
% % we are actually dealing with.
% % Clarifying the relationships between Pango Xs helps with this task.

% The Pango X lineage designations represent the community’s best attempt to
% catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
% manual, volunteer-driven process that synthesizes multiple lines of evidence
% and expert judgment rather than a single automated criterion.  As a result, the
% system provides the closest available approximation to “truth data,” but
% inevitably contains inconsistencies and over-splitting.  Within the
% \texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
% directly to recombination nodes (Type I events).  Examination of their
% genealogical relationships shows that many of these 44 lineages share common
% ancestral recombination events rather than arising independently.  In
% particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
% recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
% and XAP—share a second common ancestor that we refer to as “Xx.”  These
% clusters account for roughly half of all X designations in the dataset,
% indicating that the community process has at times produced multiple labels for
% the descendants of a single recombinant lineage.  Conversely, several [TODO 
% be more concrete]
% designated X lineages show little or no evidence of recombination in the ARG,
% suggesting that some may reflect convergent mutation or limited sampling rather
% than hybrid origin.  Together, these results provide a data-driven synthesis in
% which the 44 currently recognized recombinant lineages collapse into a smaller
% number of well-supported ancestral events, simplifying the complex picture of
% Omicron-era evolution and bringing the community nomenclature into closer
% alignment with the underlying genealogy.

% % P4. Results on global case numbers and frequency of recomb.
% % SHZ: Should we mention how the reliability of global case number data
% % and variation in sampling effort over time and across countries
% % might affect our interpretation of these findings?
% [FIXME: This is direct output from ChatGPT.  Review!]
% The dated ARG allows us to place recombination within its epidemiological
% context, revealing clear temporal structure in when recombinant lineages arise.
% The frequency of inferred recombination events follows global case numbers
% through the pandemic, with peaks during the Delta and Omicron waves when both
% infection prevalence and genetic diversity were high.  Quantitatively, the best
% single predictor of weekly recombinant counts is the product of global cases
% and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
% variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
% relationship is consistent with the mechanistic expectation that recombination
% requires both frequent co-infections (driven by high case numbers) and
% genetically distinct co-infecting viruses (driven by lineage diversity).  In
% other words, recombination becomes most detectable when epidemiological and
% evolutionary opportunity coincide.  These results imply that the emergence of
% novel recombinants is not uniform through time but concentrated into
% predictable ``hot periods'' during large, antigenically diverse outbreaks, when
% the ecological conditions for viral recombination are most favourable.

% % P5. Confirm breakpoints around Spike. 
% % SHZ: Maybe we can say it's a known (instead of robust)
% % pattern of detectable recombination in CoVs in general?
% % There are some citations for template switching hotspots in SC2.
% % Also, it's not only about selection favoring new combinations of mutations in Spike.
% % It's about combinations of mutations in, say, ORF1ab, getting selected against
% % (or just low mutation rates).
% % Maybe mention epistasis?
% [FIXME: This is directish output from ChatGPT. Review!]
% The genomic distribution of breakpoints in the ARG is highly uneven, with a
% pronounced enrichment near the $5'$ boundary of the Spike gene and across the
% S1–S2 junction.  This pattern closely mirrors that observed in the
% UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
% analyses \citep{Jackson2021} [more cites], but here it is recovered
% independently over a much longer sampling period and under a more conservative
% inference regime.  Under our more conservative estimate of the overall
% prevalence of recombination, we find that this enrichment persists despite a substantial
% reduction in the total number of inferred recombination events, indicating that
% it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
% by-product of detection sensitivity.  Across all high-confidence events, the
% median breakpoint interval length is $\sim 2.2$kb and declines with parental
% divergence, consistent with expectations from sequence-based localization
% accuracy.  These results show that even with a conservative lower bound on
% recombination frequency, the concentration of breakpoints around Spike remains
% a robust signature of coronavirus genome biology—likely reflecting a
% combination of mechanistic biases in template switching and selective
% constraints on Spike function.

% % P6. Sc2ts method is a major advance. We validate with real data
% % rather than simulations. Forward looking statement about simulations.
% % 
% % SHZ: Could at least some of this go into the Introduction?
% In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
% it is now possible to reconstruct recombination-aware genealogies at the scale
% of entire pandemics.  Although recombination is recognised as a major
% evolutionary force in many pathogens—including RNA viruses such as influenza
% and HIV, and numerous bacterial species—genealogical analyses have typically
% relied on tree-based approximations that ignore or collapse recombination.
% This simplification has limited the accuracy of downstream evolutionary and
% epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
% overcomes this barrier by directly inferring an ancestral recombination graph
% from millions of viral genomes, producing the first coherent reticulate
% genealogy for any pathogen dataset of comparable size.  The ability to
% integrate recombination into genome-wide genealogies at this scale represents a
% major conceptual step forward: it transforms recombination from a confounding
% complication into an analyzable component of pathogen evolution.

% % P7. The software ecosystem enables diverse besplike analysis.

% [TODO: This is close, needs revision]
% The practical value of the ARG is amplified by its integration into the broader
% software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
% open-source libraries provide highly efficient storage and analysis of tree
% sequences and variant data, allowing millions of genomes to be explored
% interactively within familiar environments such as Python and Jupyter
% notebooks.  Analyses that previously required bespoke phylogenetic
% pipelines—such as extracting local genealogies, summarising ancestral
% relationships, mapping mutations to branches, or tracking recombinant ancestry
% through time—can now be performed with concise, reproducible code on standard
% hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
% ARG integrates seamlessly with widely used tools in the PyData ecosystem,
% enabling users to combine genealogical information with epidemiological,
% geographic, or phenotypic metadata in the same workflow.  This interoperability
% transforms what can be done with large viral genome collections, lowering the
% barrier to sophisticated evolutionary analysis and making recombination-aware
% genealogy inference accessible to a wide community of researchers.

% % P8. Outlook

% [TODO a bit muddled, refine]
% Together, these results demonstrate that recombination in SARS-CoV-2 can be
% systematically incorporated into genealogical inference at pandemic scale. The
% \texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
% both biologically realistic and computationally tractable, showing that
% recombination can be modelled rather than ignored. Embedded within an
% interoperable software ecosystem, it provides a comprehensive record of
% SARS-CoV-2 ancestry and a practical framework for real-time,
% recombination-aware genomic surveillance—turning large sequence archives into
% dynamic evolutionary models that can be queried with concise, reproducible code. 
% More broadly, this work marks a transition from
% tree-based pandemic phylogenetics to recombination-aware genealogy inference.
% While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
% point for other recombining pathogens: the core ideas are portable,
% even though effective deployment will require pathogen-specific adjustments. 
% We anticipate that this resource will clarify the recent history of
% SARS-CoV-2 and catalyse methodological refinement and targeted biological
% studies of recombination across systems where such extensions prove feasible.

% [NOTE: This is Shing's WIP alternative]

% %%% SHZ: WIP alternative.
% We present the first pandemic-scale
% ancestral recombination graph (ARG) for SARS-CoV-2,
% reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
% We demonstrate how this ARG, for the first time,
% enables systematic identification and characterization of
% recombination events in fine-grained
% detail--in terms of their parents, breakpoints, and timing of
% emergence--providing richer and more nuanced
% genealogical, genomic, and epidemiological context
% than existing phylogenetic approaches.
% Together, sc2ts and the ARG represent a new paradigm in
% how we study the evolution of pathogens in the face of ongoing sampling.

% Accounting for recombination is a key challenge in pathogen genomics, and
% recombination has been largely ignored in analysis of viral genomes.
% Existing phylogenetic approaches for pathogens detect recombination
% \textit{post hoc} (e.g., UShER+RIPPLES),
% treat it as a nuance (e.g., XX), or
% discard its signals from downstream analyses (e.g., XX).
% Sc2ts addresses this challenge
% by using the ARG as the underlying data model,
% explicitly representing recombination events as individual nodes and
% relating them to specific parents in the ARG.
% The ARG provides a coherent framework to determine and represent
% the relationships between recombinants and their specific parents
% (which are not possible using a tree-based data model, e.g., the MAT format of UShER).
% The ARG reveals and clarifies the relationships between
% individual samples of the same recombinant origin,
% avoiding issues such as over-naming recombinants
% (e.g., the samples assigned to
% five differently labeled recombinant Pango lineages
% descending from the “Xx” recombination event in the ARG).
% Our results show how an ARG can considerably simplify
% efforts to identify and monitor emerging recombinants,
% facilitating communication of genomic surveillance results.

% % SHZ: Very WIP
% % Discuss the new preprint from Turakhia.
% The timing of recombination events is of key interest in
% pathogen genomic epidemiology.
% because this information may yield insights into
% how the emergence of a new recombinant might be
% linked with changes in public health policies.

% The overall frequency of detectable recombination events
% estimated from the sc2ts ARG is substantially lower than
% that from UShER+RIPPLES,
% largely reflecting differences in model formulation and parameter choice.
% When applied \textit{post hoc} to an UShER phylogeny of
% $\sim$1.6 million genomes sampled up to May, 2021,
% RIPPLES reported 589 unique recombination events with 43,104 descendant
% samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
% Using the same time interval, sc2ts \(k=4\) identifies
% 14 high-confidence recombination events with
% 172 descendant samples among 501,235 samples \(0.034\%\),
% a 40- to 80-fold reduction.
% Much of this difference is explained by
% the stricter parsimony penalty used here:
% increasing the penalty from p=3 (as used in the RIPPLES analysis) to
% k=4 greatly reduces false positives
% while retaining nearly all verified recombinant lineages,
% including those missed by RIPPLES at equivalent thresholds.
% These indirect comparisons suggest that
% our estimate represents a conservative lower bound.
% In future studies, dissecting the methodological differences
% between sc2ts and UShER+RIPPLES may
% inspire new ideas to develop or improve ARG building methods
% for SARS-CoV-2 (and potentially other pathogens).

% The ARG itself is a powerful resource
% to support future studies of SARS-CoV-2.
% Our results show that the ARG captures and unifies
% important features of SARS-CoV-2 evolution,
% which are shown in different studies
% using various methods, tools, and approaches.
% The ARG also compactly encodes genetic variation in
% millions of SARS-CoV-2 genomes,
% allowing for convenient and efficient reuse of
% this valuable dataset for a range of
% genetic and evolutionary analyses beyond those presented here,
% for instance, to understand
% how the variability in sequencing efforts
% across time and location affects our ability
% to reconstruct the origin of important recombinant lineages and
% to estimate when they emerge.
% Because building the ARG is computationally expensive, and
% checking and improving various aspects of its quality is laborious,
% we hope that making this resource openly available will
% save significant time and effort for other researchers,
% enabling immediate focus on analyzing the information
% contained within the COVID-19 dataset.

% % TODO: Briefly discussion software tools here or in a subsequent paragraph.

% Recent developments in ARG methods have focused almost exclusively on
% applications to human sequencing data.
% The availability of human ARGs reconstructed from large datasets
% has already had a tremendous impact on improving
% our understanding of human genetic ancestry and past demography.
% We anticipate similar advances to be made
% by adapting existing inference methods for human ARGs
% to use the sc2ts ARG for the genealogy-based study of
% natural selection or the geographic spread of SARS-CoV-2 through time.

% The unprecedented throughput and pace of genome sequencing are
% creating ample opportunity to understand the impacts of recombination
% on pathogen evolution and epidemiology.
% We hope that this study will spur
% development of ARG methods for pathogens,
% ultimately making recombination-aware genome analysis
% routine for all sorts of pathogens beyond SARS-CoV-2.

@@

% SHZ: WIP
We close by emphasizing several current limitations
and potential directions to improve sc2ts,
which we hope will also inspire future research
to develop ARG inference methods for pathogens beyond SARS-CoV-2.

We suggest several ways to extend the HMM.
First, reversion of deletions can indicate recombination,
because bases that are lost are unlikely to be regained,
except via recombination with a lineage that has those bases.
We presently treated multi-base deletions as missing data, and
added deletions to the ARG \textit{post hoc}.
Instead, by explicitly modeling deletions (and more generally insertions) in the HMM,
we should be able to recover additional signals of recombination
and reduce spurious recombinants.
Second, the current HMM assumes a simplistic mutation model
(where mutation to any other allele is equally probable) and
a uniform mutation rate along the SARS-CoV-2 genome.
However, genomic analyses have revealed
transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024} and
sequence context-dependent mutation rate variation
across the SARS-CoV-2 genome~\cite{Haddox2025}.
Using a more realistic mutation model and
accounting for mutation rate heterogeneity
could improve inference of local genetic relationships in the ARG.
Third, geography can be accounted for in the HMM,
favoring attachment of samples to geographically close lineages.
This should also improve precise identification of recombinant parents,
because parents sampled in the same region or geographically close regions
are expected to be more plausible
than parents sampled in geographically distant regions.
